Refinements for open automata

### 2023-06-19 2021-Refinement/paper/refinementarticleV2.tex
relecture sec 1 et 2

\keywords{Labelled transition systems  \and Refinement \and Composition.}
\end{abstract}
%
%
%
\section{Introduction}
%
%
%\TODO{TODO list}
%
%
%* \st{example a terminer}
%
%* choisir terminologie: refinement or simulation
%
%* \st{open automata  / Open Automata -> only use open automata or OA.}
%
%* positioning
%
%* \st{format according to conf: lncs FACS or SEFM}
%
%* global rewriting (partly done)

\TODO{intro}
The present trend in software engineering is towards component-based design. Compositional design is a highly convenient approach 
for specifying and verifying large systems. Automata are the basic formalism for this approach, but most automata definitions allow only the specifications of finite closed systems. Indeed those systems are the ones we can verify efficiently, but programming often consists in writing open systems that should be interfaced with others, and with potentially unbound behaviours (at least statically). We investigate in our works the possibility to define open symbolic systems, verify some of their properties, and use them to compose in a structured way more complex systems.

\RAB{The refinement concept plays  an important role in software engineering. In addition to helping to cope with the complexity, refinement provides a foundation for ensuring  system (software) correctness. The correctness of a system can
be established by proving, that a system refines its specification.}
%Refinement relations verify that an implementation satisfies some properties with respect to a specification. 
Refinement entails that one system  can be considered as a more precise version of the specification, featuring all the specified behaviours with more concrete details. From  formal point of view,  refinement can be defined as a mathematical relations between a specification and its implementation. It is usually defined as trace inclusion or simulation  \cite{Milner:1980,Kouchnarenko:2007}. Each  offers specific guarantees on the  behavioural properties of the specification, and it ensures that all behaviours of the specification must belong to the behaviours of the implementation. This definition, which is based on systems whose behaviour is fully defined, is not well-suited for open systems,  as it requires to reason also about unspecified behaviours.

Open automata (that we abbreviate OA) were defined as a way to provide a semantics for open parameterized hierarchical LTSs used in verification tools and called \emph{pNets}.
An OA \cite{henrio:01299562} is a classical automaton with variables and holes. Variables make automata symbolic and allow them to encode infinite state systems. Holes enable the composition of automata: an automaton with a hole is an operator that takes another automaton as parameter and reacts to the actions it emits; the composed automaton is a more precise automaton where the behaviour of one ``process parameter'' % \TODO{est-ce qu'on n'ecrirait pas ici "process parameter" pour les distinguer des variables standard ?}
% LUDO: DONE
 of the main automaton has been provided.

@@

%This article first introduces OA that were defined in  \cite{henrio:01299562}. %,  but compared to previous works, we and additionally defines the composition of OA.
Our contribution in this paper is the definition a refinement relation for OA that has the following characteristics:
\begin{itemize}
\item Classical simulation characterisation but also an additional criteria ensuring that refinement does not introduce deadlocks when following a trace from the simulated automaton.
\item Good properties relatively to composition: we prove here that filling the same hole  with the same automaton preserves the refinement relation.
\item Ability to take into account both composition and transitivity: this is a challenge because composition changes the set of holes of the OA and refinement takes into account the actions of the holes.
\end{itemize}
 The refinement relation is introduced in two steps. First we define a refinement that relates two automata with the same holes, which allows us to focus on the automaton aspect. Second we introduce a relation that relates two automata with different sets of holes, which allows us to take into account the open nature of OA, and to deal with composition. Properties of the refinement are stated and proven on the second, more general version of the relation, they are thus also valid for the first simpler refinement relation.

@@

\subsection{Preliminaries and notations}
Families of values, or equivalently maps will be noted \(x_i^{i \in I}\), \(\mset{i \mapsto x_i}{i \in I}\), or \(\mset{i \gets x_i}{i \in I}\), depending on what is more convenient (e.g. $i \gets x_i$ is used for maps that are used as substitution). Statements like  $\exists c_j^{j \in J}$ defines both $J$ and the mapping ${j \mapsto c_j}$.
%For instance \(\mpar{ax}^{x \in \setR}\) represents a scaling function, \(c^{i \in I}\) is a constant function over \(I\).
%They will be used depending on what is more convenient.
%For instance \(\mbrc{\alpha \mapsto 1, \beta \mapsto 2, \gamma \mapsto 3}\) has no simple generating expression and is better represented with the finite version of first notation.
The disjoint union on sets is noted \(\uplus\).
%\footnote{Depending on the context, \(\uplus\) notation either supposes that the sets are disjoint or constraints joined sets to be disjoints if several choices are possible}.
Disjoint union is also used on maps.
There are several ways of ensuring a union is disjoint, we will indifferently either suppose sets are disjoint or rename conflicting objects (useful for variables).
%The disjoint union of two maps \(\varphi: I \to X\) and \(\psi: J \to Y\) with \(I \cap J = \emptyset\) is noted \(\varphi \uplus \psi\) and has the following signature \(I \uplus J \to X \cup Y\).
In a formula, a quantifier followed by a finite set will be used as a shorthand for the quantification on every variable in the set:
\(\forall \mbrc{a_1, \dots, a_n}, \exists \mbrc{b_1, \dots, b_m}, P\) means \(\forall a_1, \dots, \forall a_n, \exists b_1, \dots, \exists b_m, P\).

@@

An \emph{open transition} is a tuple \nmm{\OTg} with \(s, s' \in S\) the source and target states, \(\alpha \in \actions\) the produced action, \(J' \subseteq J\) the holes involved in the transition, \(\beta_j \in \actions\) the actions of the holes, \(g \in \formulas\) the guard and \(\psi: V \to \terms\) the variable assignments.
To be well-formed, an open transition should use only variables of the automaton and variables appearing in the involved actions, formally: 
\begin{align*}
\fvars{g}&\subseteq \fvars{\alpha}\cup \bigcup_{j \in J'} \fvars{\beta_j} \cup V \\ \forall v\in V.\, \fvars{\psi(v)}&\subseteq \fvars{\alpha}\cup \bigcup_{j \in J'} \fvars{\beta_j} \cup V
\end{align*}
\end{definition}

We use two operators to access pieces of information of the OA.

%\TODO{useful or not?}
\begin{definition}[Out-transition, Transition variables]
Let \(V\) be the variable names of the considered automaton, \(T\) its transitions and \(r\) one of its states.
\(\fOT[T]{r} \subset T\) are called the out-transitions of the state \(r\).
%\(\fIT[T]{r} \subset T\) are called the in-transitions of the state \(r\).
When the transition set is clear from the context, it will be omitted.
The local variables of a transition \(\fvars{t}\) are all variables appearing in transition \(t\) except the variables of the automaton.
%\begin{align*}
% %&
%%	\fIT[T]{r} & = \mset{\OTg \in T}{s' = r} \\
%\end{align*}
%\vspace{-1cm}
\begin{mathpar}
%	\fguard{\OTg} \!=\! g \qquad
	\fOT[T]{r}  = \mset{\OTg \in T}{s = r}

@@

A pair consisting of a state and a valuation is called a \emph{configuration}.
%
%The following terminology will be used to reason on open automata
%\begin{definition}[Configuration, instantiated transition]
%An \emph{instantiated transition} of an automaton \(\OAg\) is a transition  \(t\psi\) where $t\in T$ and $\psi$ is a  well-formed substitution with $\dom(\psi)=vars(t)$. %\setminus V$.
%\end{definition}

@@

The example uses several kinds of data. Variable $\texttt{m}$ holds a message, message types do not need a concrete definition here. We additionally use arrays of messages with a syntax of the form $\texttt{M[l]}$ for array accesses; $\texttt{M}$ is an array of $\texttt{N}$ elements, from $0$ to $\texttt{N}-1$. Finally we use addition and modulo operation ($\%$) on integers.

@@

OA are partially specified automata, the partiality arises  from the holes.
A hole can be seen as a port in which we can plug an OA.
The plugging operation is called composition.
The composition of OA was already implicitly defined by the means of composition on pNets in previous work \cite{henrio:01299562} but never  formalised on OA.\footnote{The definition of composition below is a direct translation of what happens with pNets composition \cite{henrio:01299562} without  introducing pNets.}

@@

\begin{example} \label{Example:prodandcons} Fig. \ref{Fig:procandcons} shows a producer automaton  and a consumer  automaton that can be used  to fill the holes $\texttt{P}$ and $\texttt{Q}$ of \texttt{prod-cons} defined in Example~\ref{Example:fifo-system}.
 \begin{figure}[!ht]
 \centering
   \includegraphics[width=.35\textwidth]{Figures/P-proc.pdf}\hfill 
   \includegraphics[width=.45\textwidth]{Figures/Q-proc.pdf}
   \caption{(Left) A \texttt{producer}.  It produces one item at a time and pushes it.  (Right)  A \texttt{consumer}. It pops an item, does some work and pushes the result. \label{Fig:procandcons}}
\end{figure}

@@

\begin{figure}[!ht]
 \centering
   \includegraphics[width=.99\textwidth]{Figures/compos-FIFO-Pro.pdf}
   \caption{OA for \texttt{prod-cons}[\texttt{P}/\texttt{producer}]. 
   \label{Fig:ComposeOA}} 
\end{figure}
\end{example}

\subsection{Relations between open automata}
Establishing semantic equivalences and refinement relations between different OA requires  to compare their states. For this purpose, we suppose that the variables of the two OA are disjoint (a renaming of variables may have to be applied before comparing OA states).
\begin{definition}[Relation on open automata configurations] Suppose $V_1$ and $V_2$ are disjoint.
A relation on configurations of two OA \(\OAg[1]\) and \(\OAg[2]\) is a function \(R: S_1 \times S_2 \to \rformulas[V_1 \uplus V_2]\).
\end{definition}
%\TODO{C'est etrange, la formulation "the variables they refer to", comme il n'y a plus de variables d'etat... +1}
The idea is that two states are related depending on the satisfiability of the expression relying their variables i.e., if the variables of the OA verify a certain formula. 
In other words, to each pair of states is attached a boolean formula that may refer to the  variables of each of the two OA, stating whether the two states are related or not.
Additionally, we may check that initial states of the automata are related by checking that: \(\sigma_{01} \uplus \sigma_{02} \vdash R\mpar{s_{01}, s_{02}}\).

@@

>>>>

\keywords{Labelled transition systems  \and Refinement \and Composition.}
\end{abstract}
%
%
%
\section{Introduction}
%
%
%\TODO{TODO list}
%
%
%* \st{example a terminer}
%
%* choisir terminologie: refinement or simulation
%
%* \st{open automata  / Open Automata -> only use open automata or OA.}
%
%* positioning
%
%* \st{format according to conf: lncs FACS or SEFM}
%
%* global rewriting (partly done)
 Compositional design is a highly convenient approach 
for specifying and verifying large systems. 
Automata are often used as the basic formalism for this approach, but most automata definitions allow only the specifications of finite closed systems. Indeed those systems are the ones we can verify efficiently, but programming often consists in writing open systems that should be interfaced with others, and with potentially unbound behaviours (at least statically). We investigate in our works the possibility to define open symbolic systems, verify some of their properties, and use them to compose in a structured way more complex systems.

The refinement concept also plays  an important role in software engineering. In addition to helping to cope with the complexity \LUDO{of what?}, refinement provides a foundation for ensuring  system (software) correctness. The correctness of a system can
be established by proving, that a system refines its specification.
%Refinement relations verify that an implementation satisfies some properties with respect to a specification. 
Refinement entails that one system  can be considered as a more precise version of another one that is considered to be the specification. The refined model features all the specified behaviours with more concrete details. From a formal point of view,  refinement can be defined as a mathematical relations between a specification and its implementation, one frequently used relation is trace inclusion or simulation  \cite{Milner:1980,Kouchnarenko:2007}. Each \LUDO{each what?} offers specific guarantees on the  behavioural properties of the specification, and it \LUDO{what is ``it''?} ensures that all behaviours of the specification must belong to the behaviours of the implementation. This definition \LUDO{which definition?}, which is based on systems whose behaviour is fully defined, is not well-suited for open systems,  as it requires to reason also about unspecified behaviours.

Open automata (that we abbreviate OA) were defined as a way to provide a semantics for open parameterized hierarchical labelled transition systems (abbreviated LTS). They were proposed as a theoretical foundation for parametrised automatas used in verification tools and called \emph{pNets}.
An OA \cite{henrio:01299562} is similar to a classical automaton but with variables and holes. Variables make automata symbolic and allow them to encode infinite state systems. Holes enable the composition of automata: an automaton with a hole is an operator that takes another automaton as parameter and reacts to the actions it emits; the composed automaton is an automaton where the behaviour of one ``process parameter'' % 
 of the main automaton has been provided.

@@

%This article first introduces OA that were defined in  \cite{henrio:01299562}. %,  but compared to previous works, we and additionally defines the composition of OA.
Our contribution in this paper is the definition a refinement relation for OA that has the following characteristics:
\begin{itemize}
\item Classical simulation characterisation but also an additional criteria ensuring that refinement does not introduce deadlocks when following a trace from the simulated automaton.
\item Good properties relatively to composition: we prove that composition preserves the refinement relation.
\item Ability to take into account both composition and transitivity: this is a challenge because composition changes the set of holes of the OA and refinement takes into account the actions of the holes.
\end{itemize}
 The refinement relation is introduced in two steps. First we define a refinement that relates two automata with the same holes, which allows us to focus on the automaton aspect. Second we introduce a relation that relates two automata with different sets of holes, which allows us to take into account the open nature of OA, and to deal with composition. Properties of the refinement are stated and proven on the second, more general version of the relation, they are thus also valid for the first simpler refinement relation.

@@

\subsection{Preliminaries and notations}
Families of values (equivalent to maps) will be noted \(x_i^{i \in I}\), \(\mset{i \mapsto x_i}{i \in I}\), or \(\mset{i \gets x_i}{i \in I}\), depending on what is more convenient (e.g. $i \gets x_i$ is used for maps that are used as substitution). Statements like  $\exists c_j^{j \in J}$ defines both $J$ and the mapping ${j \mapsto c_j}$.
%For instance \(\mpar{ax}^{x \in \setR}\) represents a scaling function, \(c^{i \in I}\) is a constant function over \(I\).
%They will be used depending on what is more convenient.
%For instance \(\mbrc{\alpha \mapsto 1, \beta \mapsto 2, \gamma \mapsto 3}\) has no simple generating expression and is better represented with the finite version of first notation.
The disjoint union on sets is noted \(\uplus\).
%\footnote{Depending on the context, \(\uplus\) notation either supposes that the sets are disjoint or constraints joined sets to be disjoints if several choices are possible}.
Disjoint union is also used on maps.
There are several ways of ensuring a union is disjoint, we will indifferently either suppose sets are disjoint or rename conflicting objects (useful for variables).
%The disjoint union of two maps \(\varphi: I \to X\) and \(\psi: J \to Y\) with \(I \cap J = \emptyset\) is noted \(\varphi \uplus \psi\) and has the following signature \(I \uplus J \to X \cup Y\).
In a formula, a quantifier followed by a finite set will be used as a shorthand for the quantification on every variable in the set:
\(\forall \mbrc{a_1, \dots, a_n}, \exists \mbrc{b_1, \dots, b_m}, P\) means \(\forall a_1, \dots, \forall a_n, \exists b_1, \dots, \exists b_m, P\).

@@

An \emph{open transition} is a tuple \nmm{\OTg} with \(s, s' \in S\) the source and target states, \(\alpha \in \actions\) the produced action, \(J' \subseteq J\) the holes involved in the transition, \(\beta_j \in \actions\) the actions of the holes, \(g \in \formulas\) the guard and \(\psi: V \to \terms\) the variable assignments.
To be well-formed, an open transition should use only variables of the automaton and variables appearing in the involved actions, formally: 
\begin{align*}
\fvars{g}&\subseteq \fvars{\alpha}\cup \bigcup_{j \in J'} \fvars{\beta_j} \cup V \\ \forall v\in V.\, \fvars{\psi(v)}&\subseteq \fvars{\alpha}\cup \bigcup_{j \in J'} \fvars{\beta_j} \cup V
\end{align*}
\end{definition}
A pair consisting of a state and a valuation is called a \emph{configuration} (of the automaton).
We use two operators to access pieces of information of the OA.

%\TODO{useful or not?}
\begin{definition}[Out-transition, Transition variables]
Let \(V\) be the variable names of a considered automaton, \(T\) its transitions and \(r\) one of its states.
\(\fOT[T]{r} \subset T\) are called the out-transitions of the state \(r\).
%\(\fIT[T]{r} \subset T\) are called the in-transitions of the state \(r\).
When the set \(T\) is clear from the context, it will be omitted.
The local variables of a transition \(\fvars{t}\) are all variables appearing in transition \(t\) except the variables of the automaton.
%\begin{align*}
% %&
%%	\fIT[T]{r} & = \mset{\OTg \in T}{s' = r} \\
%\end{align*}
%\vspace{-1cm}
\begin{mathpar}
%	\fguard{\OTg} \!=\! g \qquad
	\fOT[T]{r}  = \mset{\OTg \in T}{s = r}

@@

%
%The following terminology will be used to reason on open automata
%\begin{definition}[Configuration, instantiated transition]
%An \emph{instantiated transition} of an automaton \(\OAg\) is a transition  \(t\psi\) where $t\in T$ and $\psi$ is a  well-formed substitution with $\dom(\psi)=vars(t)$. %\setminus V$.
%\end{definition}

@@

The example uses several kinds of data. Variable $\texttt{m}$ holds a message (we can leave the message type abstract here). We additionally use arrays of messages with a syntax of the form $\texttt{M[l]}$ for array accesses; $\texttt{M}$ is an array of $\texttt{N}$ elements, from $0$ to $\texttt{N}-1$. Finally we use addition and modulo operation ($\%$) on integers.

@@

OA are partially specified automata, the partiality arises  from the holes.
A hole can be seen as a port in which we can plug an OA.
The plugging operation is called composition.
The composition of OA was already implicitly defined by the means of composition on pNets in previous work \cite{henrio:01299562}. We provide here a (new) direct definition of composition for OA.

@@

\begin{example} \label{Example:prodandcons} Fig. \ref{Fig:procandcons} shows a producer automaton  and a consumer  automaton that can be used  to fill the holes $\texttt{P}$ and $\texttt{Q}$ of \texttt{prod-cons} defined in Example~\ref{Example:fifo-system}.
 \begin{figure}[!b]
 \centering
   \includegraphics[width=.35\textwidth]{Figures/P-proc.pdf}\hfill 
   \includegraphics[width=.45\textwidth]{Figures/Q-proc.pdf}
   \caption{(Left) A \texttt{producer}.  It produces one item at a time and pushes it.  (Right)  A \texttt{consumer}. It pops an item, does some work and pushes the result. \label{Fig:procandcons}}
\end{figure}

@@

\begin{figure}[!b]
 \centering
   \includegraphics[width=.99\textwidth]{Figures/compos-FIFO-Pro.pdf}
   \caption{OA for \texttt{prod-cons}[\texttt{P}/\texttt{producer}]. 
   \label{Fig:ComposeOA}} 
\end{figure}
\end{example}

\subsection{Relations between open automata}
Establishing semantic equivalences and refinement relations between different OA requires  to compare their states. For this purpose, we suppose that the variables of the two OA are disjoint (a renaming of variables may have to be applied before comparing OA states).
\begin{definition}[Relation on open automata configurations] Suppose $V_1$ and $V_2$ are disjoint.
A relation on configurations of two OA \(\OAg[1]\) and \(\OAg[2]\) is a function \( \mathcal{R}: S_1 \times S_2 \to \rformulas[V_1 \uplus V_2]\).
\end{definition}
%\TODO{C'est etrange, la formulation "the variables they refer to", comme il n'y a plus de variables d'etat... +1}
The idea is that two states are related depending on the satisfiability of the expression relying their variables i.e., if the variables of the OA verify a certain formula. 
In other words, to each pair of states is attached a boolean formula that may refer to the  variables of each of the two OA, stating whether the two states are related or not.
Additionally, we say that the relation $ \mathcal{R}$ \emph{relates the initial states} of the automata if: \(\sigma_{01} \uplus \sigma_{02} \vdash  \mathcal{R}\mpar{s_{01}, s_{02}}\).

@@

Two automata are bisimilar if there exists a strong FH-bisimulation $ \mathcal{R}$ that relates their initial states.
